---
title: "WiFi Fingerprinting"
output:
  github_document:
    toc: true
---
# Introduction
Many real world applications rely on a user's location to provide optimal service. Thanks to the inclusion of GPS sensors in mobile devices, outdoor localization can be pinpointed very accurately. However, indoor localization remains a problem due to the loss of GPS signal in indoor environments. Wi-Fi positioning system is a geolocation system that uses the characteristics of nearby Wi-Fi hotspots and other wireless access points to discover where a device is located.

This analysis uses the [UJIIndoorLoc Data Set](http://archive.ics.uci.edu/ml/datasets/UJIIndoorLoc) to develop models capable of reliably locating the position of a user's device using signals from wireless access point (WAPs). The data set contains 520 WAPs as well as user data such as longitude, latitude, user ID, phone ID, and building-specific location values. After initial pre-processing I combined the dependent variables into one value, and then proceeded to modeling with encouraging results. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r, message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(tidyr)
```
# Importing the data
```{r}
## Download, unzip, and import the data
temp <- tempfile()
download.file("http://archive.ics.uci.edu/ml/machine-learning-databases/00310/UJIndoorLoc.zip",temp)
training <- read.csv(unz(temp, "UJIndoorLoc/trainingData.csv"))
validation <- read.csv(unz(temp, "UJIndoorLoc/validationData.csv"))
unlink(temp)
```
# Preprocessing
According to the dataset description there are 529 attributes. The first step is to identify which of these features are not necessary. I will use the nearZeroVar function in the caret package to accomplish this, and eliminate redundant attributes.
```{r}
## Create a working copy of the data 
workData<-training

## Remove zero variance features
zeroVarData <- nearZeroVar(training, saveMetrics = TRUE)
zeroVarData1 <- which(zeroVarData$zeroVar == T)

## Rewrite working copy
workData <- training[,-(zeroVarData1)]
```
Most of the attributes are WiFi fingerprints listed as integer values. The final 9 attributes contain other useful attributes including our dependent variables.
```{r}
str(workData[,466:474])
```
Some of these attributes are not needed for this evaluation. Others need to be converted to factors.
```{r}
## Set datatypes
workData$FLOOR<-as.factor(workData$FLOOR)
workData$BUILDINGID<-as.factor(workData$BUILDINGID)
workData$SPACEID<-as.factor(workData$SPACEID)
workData$RELATIVEPOSITION<-as.factor(workData$RELATIVEPOSITION)

## Remove uneeded features
workData$USERID<-NULL
workData$PHONEID<-NULL
workData$TIMESTAMP<-NULL
```
Because there are more than one attributes related to location in the dataset, there is a need to combine them into one depednent variable. 
```{r}
## Combine dependent variables
uniteData <- workData %>% unite("LOCATION", "BUILDINGID", "FLOOR", "SPACEID", remove = TRUE)
uniteData$RELATIVEPOSITION<-NULL
```
# Modeling
Three models were created and run on the dataset: native bayes, k nearest neighbor, and random forest. The CARET package is extremely useful for training and testing these models.
```{r}
## Split data into training and test sets and generate resampling procedure using 10-fold cross validation
set.seed(23)
inTraining <- createDataPartition(uniteData$LOCATION, p=.75, list = FALSE)
training <- uniteData[inTraining, ]
testing <- uniteData[-inTraining, ]
fitControl <- trainControl(method="repeatedcv", number = 10, repeats = 1)
```
```{r, cache=TRUE}
## Train the models
knnFitr1 <- train(LOCATION~., data = training, method = "kknn", trControl=fitControl)
knnFitr1
nbFit1 <- train(LOCATION~., data = training, method = "naive_bayes", trControl=fitControl)
nbFit1
rfFit1 <- train(LOCATION~., data = training, method = "rf", trControl=fitControl, tuneLength=1)
rfFit1
```
```{r}
## Compare all models using resampling
ModelData <- resamples(list(KNN = knnFitr1, NB = nbFit1, RF = rfFit1))
summary(ModelData)
```
The random forest model generates a high accuracy and kappa value. Using this model indoor locationing can be achieved with confidence.
